---
title: "Forecast performance"
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r setup}
library(ggiraph)
library(dplyr)
library(ggplot2)
library(glue)
source("R/plot-utils.R")
#source("../R/ignore_sigpipes.R")
#ignore_sigpipe()

aquatics_focal_sites <- c("USGS-05553700", "USGS-01463500")
```

This page visualizes the forecasts and forecast performance for the focal target variables.

## Most recent forecasts {#sec-performance}

Only the top performing models from the last 30 days are shown.

```{r echo = FALSE}
### GET FORECASTS
reference_datetimes <- arrow::open_dataset("../cache/summaries") |> 
  dplyr::summarize(reference_datetime_max = max(reference_datetime), .by = "variable") |>  
  dplyr::collect() |> 
  group_by(variable) |>
  dplyr::mutate(reference_datetime_max = min(c(reference_datetime_max, Sys.Date() - lubridate::days(1))))

config <- yaml::read_yaml("../challenge_configuration.yaml")
sites <- readr::read_csv(paste0("../", config$site_table), show_col_types = FALSE)

df_P1D <- arrow::open_dataset("../cache/summaries/duration=P1D") |> 
  left_join(reference_datetimes, by = "variable") |> 
  filter(reference_datetime == reference_datetime_max) |> 
  left_join(sites, by = "site_id") |> 
  filter(site_id %in% sites$site_id) |> 
  mutate(reference_datetime = lubridate::as_datetime(reference_datetime),
         datetime = lubridate::as_datetime(datetime)) |> 
  filter(lubridate::as_date(datetime) > lubridate::as_date(reference_datetime)) |> 
  collect()

```

```{r}
cutoff <- Sys.Date() - lubridate::days(30)
df_P1D_scores <- arrow::open_dataset("../cache/scores/duration=P1D") |> 
  left_join(sites, by = "site_id") |> 
  mutate(reference_datetime = lubridate::as_datetime(reference_datetime),
         datetime = lubridate::as_datetime(datetime)) |>
  filter(reference_datetime > cutoff) |> 
  collect() 


cutoff <- Sys.Date() - lubridate::days(365)


ref <- Sys.Date() - lubridate::days(30)

ref_P1D <- min(c(Sys.Date() - lubridate::days(30), 
             lubridate::as_date(df_P1D$reference_datetime)))


#n_data <- 10
#who <- combined |> filter(!is.na(observation)) |> summarise(has_data = max(reference_datetime)) |> collect()
#ref <- as.character ( as.Date(who$has_data[[1]]) - n_data )
ex_P1D <- df_P1D_scores |> 
  mutate(min_reference_datetime = min(reference_datetime)) |> 
  filter(reference_datetime == min_reference_datetime)

```

```{r}
#Best models

best_P1D_scores <- df_P1D_scores |> 
  summarise(score = mean(crps, na.rm = TRUE), .by = c("model_id","variable")) |> 
  arrange(variable, score) |> 
  group_by(variable) |> 
  slice(1:5)

```
Forecasts submitted on `r max(lubridate::as_date(df_P1D$reference_datetime))`


### River Chlorophyll-a

Forecast summaries are available [here](https://radiantearth.github.io/stac-browser/#/external/raw.githubusercontent.com/eco4cast/usgsrc4cast-ci/main/catalog/summaries/aquatics/Daily_Chlorophyll_a/collection.json)

```{r}
best_models <- best_P1D_scores |> filter(variable == "chla") |> pull(model_id)

df_P1D |> 
  filter(variable == c("chla"),
         model_id %in% best_models, 
         site_id %in% aquatics_focal_sites
         ) |> 
  mutate(observation = as.numeric(NA)) |> 
  forecast_plots()

```

:::

## Forecast analysis

Below are forecasts submitted 30 days ago and include the observations used to evaluate them. Mouse over to see the team id, scroll to zoom.  Only the top five performing models are shown. Information on how to access the scores can be found in our [catalog](https://radiantearth.github.io/stac-browser/#/external/raw.githubusercontent.com/eco4cast/usgsrc4cast-ci/main/catalog/scores/collection.json)

::: panel-tabset

### River Chlorophyll-a

```{r}
best_models <- best_P1D_scores |> filter(variable == "chla") |> pull(model_id)


ex_P1D |> 
  filter(variable == c("chla"),
         model_id %in% best_models, 
         site_id %in% aquatics_focal_sites
         ) |> 
  forecast_plots()

```

:::

## Aggregated scores

Average skill scores of each model across all sites.\

Scores are shown by reference date and forecast horizon (in days).\

Scores are averaged across all submissions of the model with a given horizon or a given `reference_datetime` using submissions made since `r cutoff`.\

Learn about the continous ranked probablity score [here](https://projects.ecoforecast.org/neon4cast-docs/Evaluation.html)

::: panel-tabset

### Aquatics: chrophyll-a

```{r}
leaderboard_plots(df_P1D_scores, "chla")
```

:::
